Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2015-04-13 Rmarkdown/biology_background.Rmd
Fix citations

[^other_DNAm]: Two additional examples of DNA methylation are N6-methyladenine (m6A) and N4-methylcytosine (m4C). N6-methyladenine is a methylated form of adenine, which is found in mRNA \citep[e.g.,]{Adams:1975ki} and DNA, although the latter only in bacterial DNA \citep{Ratel:2006dv}. N4-methylcytosine has also been detected in bacterial DNA \citep[e.g.,]{Ehrlich:1985bv, Ratel:2006dv}.

@@

A frequently used analogy when describing epigentic marks refers to 'writers', 'readers' and 'erasers' \citep[e.g.,]{Moore:2013in}. In the case of DNA methylation, writers catalyse the methyl group onto the DNA, readers recognise methylated DNA and erasers remove the methyl group from the DNA.

@@

Almost all assays of DNA methylation measure a population average from a pool of hundreds or thousands of cells. For a diploid organism, this is an average over several distinct levels: the two DNA strands, the two homologous chromosomes within a diploid cell and the hundreds or thousand of cells used in the assay. Hundreds or thousands of cells are required in order to have sufficient material as input for the assay. Assays that require only a single cell as input do exist \citep[e.g.,]{Smallwood:2014kn, Guo:2013ih} but are still in development and not yet in widespread use.

>>>>

[^other_DNAm]: Two additional examples of DNA methylation are N6-methyladenine (m6A) and N4-methylcytosine (m4C). N6-methyladenine is a methylated form of adenine, which is found in mRNA \citep[e.g.,][]{Adams:1975ki} and DNA, although the latter only in bacterial DNA \citep{Ratel:2006dv}. N4-methylcytosine has also been detected in bacterial DNA \citep[e.g.,][]{Ehrlich:1985bv, Ratel:2006dv}.

@@

A frequently used analogy when describing epigentic marks refers to 'writers', 'readers' and 'erasers' \citep[e.g.,][]{Moore:2013in}. In the case of DNA methylation, writers catalyse the methyl group onto the DNA, readers recognise methylated DNA and erasers remove the methyl group from the DNA.

@@

Almost all assays of DNA methylation measure a population average from a pool of hundreds or thousands of cells. For a diploid organism, this is an average over several distinct levels: the two DNA strands, the two homologous chromosomes within a diploid cell and the hundreds or thousand of cells used in the assay. Hundreds or thousands of cells are required in order to have sufficient material as input for the assay. Assays that require only a single cell as input do exist \citep[e.g.,][]{Smallwood:2014kn, Guo:2013ih} but are still in development and not yet in widespread use.

### 2015-04-13 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Fix citations

Read mapping is complicated by the bisulfite-treatment of the DNA. Following bisulfite-treatment, the DNA fragments are mostly composed of three bases rather than four, which means there are many more sequence mismatches between a read and its true mapping location. Simply using standard read mapping software and allowing for more mismatches would result in many reads mapping to multiple locations in the reference genome. Instead, a field of read mapping software dedicated to bisulfite-sequencing data has developed. Several review articles have summarised and compared the various approaches \citep[e.g.,]{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.

@@

Methods to estimate the cellular heterogeneity bias and adjust for it are available \citep[e.g.,]{Jaffe:2014gy,Houseman:2014ds,Zou:2014gc}, although they have mostly been applied to DNA methylation arrays and not bisulfite-sequencing data. This is not to say that these problems don't exist for sequencing data, merely that these have not been as well-explored.

>>>>

Read mapping is complicated by the bisulfite-treatment of the DNA. Following bisulfite-treatment, the DNA fragments are mostly composed of three bases rather than four, which means there are many more sequence mismatches between a read and its true mapping location. Simply using standard read mapping software and allowing for more mismatches would result in many reads mapping to multiple locations in the reference genome. Instead, a field of read mapping software dedicated to bisulfite-sequencing data has developed. Several review articles have summarised and compared the various approaches \citep[e.g.,][]{Chatterjee:2012ft, Krueger:2012ks, KundeRamamoorthy:2014fj}.

@@

Methods to estimate the cellular heterogeneity bias and adjust for it are available \citep[e.g.,][]{Jaffe:2014gy,Houseman:2014ds,Zou:2014gc}, although they have mostly been applied to DNA methylation arrays and not bisulfite-sequencing data. This is not to say that these problems don't exist for sequencing data, merely that these have not been as well-explored.
